---
title: "Harvesting UCEs from Genomes"
author: "Megan Ewing"
date: "2025-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is following [phyluce tutorial III: harvesting UCE loci from genomes](https://phyluce.readthedocs.io/en/latest/tutorials/tutorial-3.html).

Important folder paths:

-   genomes: `../data/genomes_refseq/ncbi_dataset/data/`

Necessary packages: [fatotwobit](https://anaconda.org/bioconda/ucsc-fatotwobit) , [twobitinfo](https://anaconda.org/bioconda/ucsc-twobitinfo), and [twobittofa](https://anaconda.org/bioconda/ucsc-twobittofa) (which i have in an enviornment 'twoBit' for):

```{bash}

### to open the twoBit environment where the packages live: 
# 
# conda activate twoBit
# 
### if conda activate gives an error like command not found, you'll need to run:
# 
# eval "$(/opt/anaconda/anaconda3/bin/conda shell.bash hook)"
# 
### and then 
# 
# conda activate twoBit
#
#
### don't forget to `conda deactivate` at end of session!
```

Need to move the genomes to the phyluce data folder as 2bits -- do this as loop. also get info

### TwoBit Genomes

```{bash}

# Input and output base directories
INPUT_BASE="../data/genomes_refseq/ncbi_dataset/data"
OUTPUT_BASE="../data/phyluce"

# Loop over each accession-named folder in the input directory
for acc_dir in "$INPUT_BASE"/*; do
  # Get the accession ID (folder name)
  accession=$(basename "$acc_dir")

  # Define input and output file paths
  fna_file="$acc_dir/cds_from_genomic.fna"
  out_dir="$OUTPUT_BASE/$accession"
  two_bit_file="$out_dir/${accession}.2bit"
  sizes_file="$out_dir/sizes.tab"

  # Check if .fna file exists
  if [[ ! -f "$fna_file" ]]; then
    echo "No .fna file found for $accession, skipping."
    continue
  fi

  # Create output directory if needed
  mkdir -p "$out_dir"

  # Convert .fna to .2bit
  echo "Converting $accession -> .2bit"
  /opt/anaconda/anaconda3/envs/twoBit/bin/faToTwoBit "$fna_file" "$two_bit_file"

  # Get sizes info
  /opt/anaconda/anaconda3/envs/twoBit/bin/twoBitInfo "$two_bit_file" "$sizes_file"

  # Preview sizes file
  echo "Preview of $sizes_file:"
  head -n 5 "$sizes_file"
  echo "-------------------------------------"
done


```

### Bait Sequences for UCE

Bait sequences were provided courtesy of Yi-Xuan Li, of Hong Kong Baptist University, from [Yi-Xuan et al. (2024)](https://academic.oup.com/sysbio/article-abstract/74/1/16/7758682)

### Align the probes to the genomes

"Now, we need to think of some name for the database to create (here \`tutorial3.sqlite\`), the name of an output in which to store the \`lastz\` search results, the path to the genome sequences, the name of the genome sequences, the path to the probe file, and a number of compute cores to use:" (from phyluce tutorial)

currently, the folders and files within are `GCF_#########.#`, but the `.#` will confuse phyluce, so let's get rid of those.

```{r}
library(fs)
library(stringr)

base_dir <- "../data/phyluce"

# rename directories
dirs <- dir_ls(base_dir, type = "directory", recurse = FALSE, regexp = "^.*/GCF_.*\\.\\d+$")
for (dir in dirs) {
  dir_name <- path_file(dir)
  new_dir_name <- str_replace(dir_name, "\\.\\d+$", "")
  new_dir_path <- path(path_dir(dir), new_dir_name)
  file_move(dir, new_dir_path)  # <- Corrected line
  cat("Renamed dir:", dir, "->", new_dir_path, "\n")
}


# rename 2bit files
# List all .2bit files under subdirectories (recursive = TRUE)
bit_files <- list.files(path = base_dir, pattern = "^GCF_\\d+\\.\\d+\\.2bit$", 
                        recursive = TRUE, full.names = TRUE)

for (old_path in bit_files) {
  # Extract current filename
  filename <- basename(old_path)
  
  # Remove the dot-version (e.g., .1) from filename
  new_filename <- sub("(GCF_\\d+)\\.\\d+\\.2bit", "\\1.2bit", filename)
  
  if (new_filename != filename) {
    # Build new path in same directory
    dir_path <- dirname(old_path)
    new_path <- file.path(dir_path, new_filename)
    
    # Rename the file
    file.rename(old_path, new_path)
    cat("Renamed:", old_path, "->", new_path, "\n")
  }
}
```

```{bash}
# list directory contents to check
# reason for the extra ls commands is to have it print in a way that can just be copied and pasted as the scaffold in the probe align chunk
ls -m ../data/phyluce/ | tr -d ','
```

We also need to adjust the sequence names within the .2bit files. now, this may seem a little two-steps-forward-one-step-back, but to do this we'll convert the .2bit files back to fasta, change the sequence names, and then convert back to .2bit and delete the intermediate files. The reason we didn't just change the fastas in the first place, is because we want to keep those exactly how they came from ncbi. we just want some change in the .2bit sequence names so that we can run lastz without the names getting overly truncated.

```{bash}

# Base directory where .2bit files live
BASE_DIR="../data/phyluce"

# Loop through all .2bit files
find "$BASE_DIR" -name "*.2bit" | while read -r twoBitFile; do
    # Extract directory and base name
    dir=$(dirname "$twoBitFile")
    base=$(basename "$twoBitFile" .2bit)

    echo "Processing: $twoBitFile"

    # Convert to .fasta
    /opt/anaconda/anaconda3/envs/twoBit/bin/twoBitToFa "$twoBitFile" "$dir/${base}.fasta"

    # Remove lcl| from headers
    sed 's/^>lcl|/>/' "$dir/${base}.fasta" > "$dir/${base}_clean.fasta"

    # Convert back to .2bit
    /opt/anaconda/anaconda3/envs/twoBit/bin/faToTwoBit "$dir/${base}_clean.fasta" "$dir/${base}_clean.2bit"

    echo "Cleaned .2bit written: $dir/${base}_clean.2bit"
done

# Replace original .2bit files with cleaned versions
find "$BASE_DIR" -name "*_clean.2bit" | while read -r cleanBit; do
    original="${cleanBit%_clean.2bit}.2bit"
    mv "$cleanBit" "$original"
done

# Optionally remove intermediate FASTAs
find "$BASE_DIR" -name "*.fasta" -delete

```

you'll need to ensure [phyluce](https://github.com/faircloth-lab/phyluce/releases/tag/v1.7.3) is installed for this

phyluce installed mamba environment: `/home/shared/8TB_HDD_02/mewing0/miniforge3/envs/phyluce/`

```{bash}
## info to initialize mamba / phyluce in terminal : 
#
# conda create -n phyluce-py3 python=3.8 phyluce -c bioconda -c conda-forge
#
# eval "$(/home/shared/8TB_HDD_02/mewing0/miniforge3/bin/conda shell.bash hook)"
# eval "$(mamba shell hook --shell bash)"
#
## activates base 
# mamba activate
#
## activates environment i created
# mamba activate phyluce 
```

```{bash}
# for v1 baits

cd ../data/phyluce/

/home/shared/8TB_HDD_02/mewing0/miniforge3/bin/conda run -n phyluce \
/home/shared/8TB_HDD_02/mewing0/miniforge3/envs/phyluce-1.7.3/bin/phyluce_probe_run_multiple_lastzs_sqlite \
        --db uce_frombaits_v1 \
        --output v1-genome-lastz \
        --scaffoldlist GCF_002022765 GCF_002113885 GCF_020536995 GCF_021730395 GCF_021869535 GCF_025612915 GCF_026571515 GCF_026914265 GCF_031769215 GCF_032062105 GCF_033153115 GCF_036588685 GCF_041381155 GCF_902652985 GCF_947568905 GCF_963676685 GCF_963853765 \
        --genome-base-path ./ \
        --probefile Baits_Bivalve_v1.fasta \
        --cores 10 \
        2> phyluce.log
        
```

```{bash}

# for v2 baits

cd ../data/phyluce/

/home/shared/8TB_HDD_02/mewing0/miniforge3/bin/conda run -n phyluce \
/home/shared/8TB_HDD_02/mewing0/miniforge3/envs/phyluce-1.7.3/bin/phyluce_probe_run_multiple_lastzs_sqlite \
        --db uce_frombaits_v2 \
        --output v2-genome-lastz \
        --scaffoldlist GCF_002022765 GCF_002113885 GCF_020536995 GCF_021730395 GCF_021869535 GCF_025612915 GCF_026571515 GCF_026914265 GCF_031769215 GCF_032062105 GCF_033153115 GCF_036588685 GCF_041381155 GCF_902652985 GCF_947568905 GCF_963676685 GCF_963853765 \
        --genome-base-path ./ \
        --probefile Baits_Bivalve_v2.fasta \
        --cores 10 \
        2> phyluce_v2.log

```

### Extracting FASTA sequence matching UCE loci from genome sequences

need to create a configuration file that tells the program where to find each genome

```{r}
library(fs)
library(stringr)

# Set paths
base_dir <- path_expand("~/bivalve_gametogenesis/data/phyluce")
output_file <- path_expand("~/bivalve_gametogenesis/data/phyluce/genomes.conf")  

# List all .2bit files recursively in GCF_* directories
bit_files <- dir_ls(base_dir, recurse = TRUE, glob = "*.2bit")

# Filter for GCF-named .2bit files
bit_files <- bit_files[grepl("GCF_\\d+/", bit_files)]

# Build config lines
config_lines <- c("[scaffolds]")

for (file_path in bit_files) {
  file_name <- path_file(file_path)
  gcf_id <- str_match(file_name, "^(GCF_\\d+)")[,2]
  full_path <- path_expand(file_path)
  config_lines <- c(config_lines, paste0(gcf_id, ":", full_path))
}

# Write to genomes.conf
writeLines(config_lines, output_file)
```

```{bash}
head -2 ../data/phyluce/genomes.conf
```

```{bash}

# for v1

cd ../data/phyluce/

/home/shared/8TB_HDD_02/mewing0/miniforge3/bin/conda run -n phyluce \
/home/shared/8TB_HDD_02/mewing0/miniforge3/envs/phyluce-1.7.3/bin/phyluce_probe_slice_sequence_from_genomes \
        --lastz v1-genome-lastz \
        --conf genomes.conf \
        --flank 500 \
        --name-pattern "Baits_Bivalve_v1.fasta_v_{}.lastz.clean" \
        --output v1-genome-fasta
```

```{bash}
# for v2

cd ../data/phyluce/

/home/shared/8TB_HDD_02/mewing0/miniforge3/bin/conda run -n phyluce \
/home/shared/8TB_HDD_02/mewing0/miniforge3/envs/phyluce-1.7.3/bin/phyluce_probe_slice_sequence_from_genomes \
        --lastz v2-genome-lastz \
        --conf genomes.conf \
        --flank 500 \
        --name-pattern "Baits_Bivalve_v2.fasta_v_{}.lastz.clean" \
        --output v2-genome-fasta \

```

below creates a summary table showing how many sequences hit for each step. notes that the fasta sequence \# may be lower than the lastz \# due to sequences getting cutoff in the flanking region

```{r}
library(tidyverse)

# Define directories
base_dir <- "../data/phyluce"
paths <- list(
  v1_lastz = file.path(base_dir, "v1-genome-lastz"),
  v2_lastz = file.path(base_dir, "v2-genome-lastz"),
  v1_fasta = file.path(base_dir, "v1-genome-fasta"),
  v2_fasta = file.path(base_dir, "v2-genome-fasta")
)

# Function to count lines (lastz) or headers (fasta)
count_lastz <- function(file) {
  # Count non-empty lines = one match per line
  length(readLines(file))
}

count_fasta <- function(file) {
  # Count how many lines start with >
  sum(readLines(file) %>% str_starts(">"))
}

# Parse sample ID from filenames
parse_sample_lastz <- function(filename) {
  str_extract(filename, "(?<=_v_GCF_)\\d+")
}

parse_sample_fasta <- function(filename) {
  str_extract(tolower(filename), "(?<=gcf_)\\d+")
}

# Collect counts
collect_counts <- function(folder, type) {
  files <- list.files(folder, full.names = TRUE)
  if (length(files) == 0) return(data.frame())
  
  tibble(
    file = files,
    sample = case_when(
      str_detect(type, "lastz") ~ paste0("GCF_", parse_sample_lastz(basename(files))),
      str_detect(type, "fasta") ~ paste0("GCF_", parse_sample_fasta(basename(files)))
    ),
    count = if (str_detect(type, "lastz")) {
      map_int(files, count_lastz)
    } else {
      map_int(files, count_fasta)
    },
    source = type
  )
}

# Run for all sources
all_counts <- map2_dfr(paths, names(paths), collect_counts)

# Reshape to wide format
df_summary <- all_counts %>%
  select(sample, source, count) %>%
  pivot_wider(names_from = source, values_from = count, values_fill = 0) %>%
  arrange(sample)

print(df_summary)


```
