---
title: "Identifying taxa coverage for each Repro gene"
author: "Megan Ewing"
date: "2025-09-05"
output: html_document
---

Here I seek to generate a fasta file for each 'gamete generation' gene, and pull the associated sequences from each bivalve genome. These will be used to generate gene trees to compare to the consensus tree to when calculating robinson-foulds distance.

first need

-   a gene list for each species that has the SPID & corresponding sequence number for that genome (done in `02-blastoncds` )

-   list of all of the unique SPIDs contained within all gene lists combined

-   configuration file with taxa names

we can then

### all unique SPIDs contained within all gene lists combined (with counts for how many species they are in)

```{r}
# get all files with full paths
files <- list.files("../output/", pattern = "^genelist_.*\\.csv$", full.names = TRUE)

# initialize list to store SPIDs by species
spid_lists <- list()

for (f in files) {
  # extract species name from the filename
  species <- sub("^genelist_(.*)\\.csv$", "\\1", basename(f))
  
  # read in gene lists
  df <- read.csv(f, stringsAsFactors = FALSE)
  
  #get unique SPIDs from this file
  spids <- unique(df$annot_w_GO.SPID)
  
  # store them
  spid_lists[[species]] <- spids
}

#cCombine into one vector w/ species labels
all_spids <- stack(spid_lists)

# count in how many unique species each SPID appears
library(dplyr)
spid_counts <- all_spids %>%
  distinct(values, ind) %>%           # remove duplicates within a species
  count(values, name = "species_count") %>%
  rename(SPID = values)

# preview
head(spid_counts)

```

```{bash}

mkdir ../output/extractsequences/
```

```{r}

# write spid list to csv
write.csv(spid_counts, "../output/extractsequences/uniqueSPIDlistandcounts.csv")
```

### 

```{r}
# Load required libraries
# Biostrings gives functions for handling DNA/RNA/AA sequences
library(Biostrings)

# Read in the master list of SwissProt IDs (~3500)
swissprot_ids <- readLines("swissprot_ids.txt")

# Get a list of all CDS fasta files (one per species/taxon)
cds_files <- list.files("cds_fastas", pattern="_cds.fasta$", full.names=TRUE)

# Loop through each SwissProt ID
for (sp_id in swissprot_ids) {
  
  # Initialize an empty DNAStringSet to hold all sequences for this SwissProt ID
  records <- DNAStringSet()
  
  # Loop through each species’ CDS fasta
  for (cdsfile in cds_files) {
    # Extract species/taxon name from file name (remove "_cds.fasta")
    taxon <- sub("_cds.fasta$", "", basename(cdsfile))
    
    # Construct path to the gene map file for this taxon
    mapfile <- file.path("gene_maps", paste0(taxon, "_gene_map.tsv"))
    
    # Load mapping file (must have at least two columns: seq_id and swissprot_id)
    map <- read.table(mapfile, sep="\t", header=TRUE, stringsAsFactors=FALSE)
    
    # Check if this SwissProt ID is present in the mapping
    if (sp_id %in% map$swissprot_id) {
      # Get all sequence IDs in this taxon that match the SwissProt ID
      seq_ids <- map$seq_id[map$swissprot_id == sp_id]
      
      # Load the CDS fasta file as a DNAStringSet
      cds <- readDNAStringSet(cdsfile)
      
      # Keep only those sequence IDs that are both in the mapping and fasta
      keep <- intersect(seq_ids, names(cds))
      
      if (length(keep) > 0) {
        # Extract the matching sequences
        cds_sub <- cds[keep]
        
        # Rename fasta headers as "species|seq_id"
        names(cds_sub) <- paste0(taxon, "|", keep)
        
        # Add them to the running list of records for this SwissProt ID
        records <- c(records, cds_sub)
      }
    }
  }
  
  # If any sequences were found, write them to a fasta file
  if (length(records) > 0) {
    writeXStringSet(
      records, 
      filepath=file.path("swissprot_fastas", paste0(sp_id, ".fasta"))
    )
  }
}

```
